---
title: "ENMeval Vignette"
author: "Robert Muscarella & Jamie M. Kass"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ENMeval Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

`ENMeval` is an R package that performs automated runs and evaluations of ecological niche models, and currently only implements Maxent. `ENMeval` was made for those who want to "tune" their models to maximize predictive ability and avoid overfitting, or in other words, optimizing the balance between goodness-of-fit and model complexity. The primary function `ENMevaluate` does all the heavy lifting and returns a table of evaluation statistics, and for each setting combination (here, colloquially: runs), a model object and prediction raster. There are also parameter options for calculating niche overlap between predictions, running in parallel to speed up computation, and more.

## Demo

We're going to start with an occurrence dataset for *Procyon lotor*, the North American raccoon, from which we extract the coordinates.

```{r}
species.df <- read.csv('vignette_data.csv', header=TRUE)  # read in the occurrence data
head(species.df)  # examine the first six rows of the occurrence data 
species.coord <- species.df[,c(3,2)]   # isolate the coordinates in order: longitude, latitude
head(species.coord)  # examine the first six rows of the coordinate data
```

Next, let's get some climate data from WorldClim. We'll download the 19 bioclim variables at 10 arcmin resolution, which is about 11.5 km across at the equator. Now's also a good time to load the package, as it includes all the downstream dependencies (`raster`, `dismo`, etc.).

```{r, warning=FALSE, message=FALSE}
library(ENMeval)  
preds <- getData('worldclim', var='bio', res=10)  # download the global bioclim dataset at 10arcmin resolution
preds  # preds is a RasterStack, which is basically a list of rasters or grids
preds[[c(3,5)]]  # double brackets allow you to subset and isolate certain layers
```

Next, we will specify the study extent by cropping (or "clipping" in ArcGIS terms) our predictor variable rasters. These cropped rasters will be used to both delineate the area from which the background points are drawn, and project our models onto to generate prediction rasters.

```{r, message=FALSE, fig.width=7, fig.height=5}
na.extent <- extent(c(-130,-50,25,45))  # rough bounding box for U.S. including most occurrence localities
preds.ext <- crop(preds, na.extent)  # crop predictor variables by this extent
backg <- randomPoints(preds.ext[[1]], n=10000)  # generate 10,000 random backround points across this extent (only one per cell centroid)
plot(preds.ext[[1]], ext=c(-100,-90,30,35))  # zoom in a bit to annual mean temp (bio1)
points(backg)  # notice how we have pretty good coverage
```

We'll now take a look at the different partitioning methods offered by `ENMeval`.

```{r}

```

We will now perform model tuning using a variety of complexity combinations. Next, we will isolate the results table from the ENMevaluate object we produce, and using this table determine which model is the best approximation of fitting to the occurrence data without being overly complex. There are a variety of selection-techniques available, but for simplicity, here we will use AICc. We want the model with the lowest AICc score, therefore, we are interested in the model with a delta.AICc of 0. 

```{r}
models <- ENMevaluate(occ = species.coord, env = preds.ext, 
                      method = 'block', categoricals = NULL, 
                      fc = c("L", 'LQ', "H", "LQH"), 
                      bg.coords = backg, RMvalues = seq(1, 5, 0.5), 
                      parallel = TRUE, rasterPreds = TRUE)
results.table<-models@results #Isolate the results table 
results.table #Look at the tuning results. Here, you will find information about omission rates, AUC and AICc scores
best.mod.AICc<- mod.table[mod.table[14]==0][c(1,16)] #Find the row of the results table that has a delta.AICc of 0.
```





## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
