---
title: "ENMeval Vignette"
author: "Robert Muscarella & Jamie M. Kass"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ENMeval Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

`ENMeval` is an R package that performs automated runs and evaluations of ecological niche models, and currently only implements Maxent. `ENMeval` was made for those who want to "tune" their models to maximize predictive ability and avoid overfitting, or in other words, optimizing the balance between goodness-of-fit and model complexity. The primary function `ENMevaluate` does all the heavy lifting and returns a table of evaluation statistics, and for each setting combination (here, colloquially: runs), a model object and prediction raster. There are also parameter options for calculating niche overlap between predictions, running in parallel to speed up computation, and more.

## Demo

We're going to start with the (insert name) dataset, from which we extract the coordinates.

```{r}
species.df <- read.csv('vignette_data.csv', header=TRUE) #Read in the occurrence data
head(species.df) #Examine the first six rows of the occurrence data 
species.coord <- species.df[,c(3,2)]  #Isolate the coordinates in order: longitude, latitude
head(species.coord) #Examine the first six rows of the coordinate data
```

Next, let's get some climate data from WorldClim. We'll download the 19 bioclim variables at 10 arcmin resolution, which is about 11.5 km across at the equator. Now's also a good time to load the package, as it includes all the downstream dependencies (`raster`, `dismo`, etc.).

```{r, warning=FALSE, message=FALSE}
library(ENMeval)  
preds <- getData('worldclim', var='bio', res=10) #Download the global bioclim dataset at 10arcmin resolution
preds  # preds is a RasterStack, which is basically a list of rasters or grids
preds[[c(3,5)]]  # double brackets allow you to subset and isolate certain layers
```

An essential step before proceeding much further is to specify the study extent, which will be used 1) to delineate from where the background points are drawn, and 2) as the data we'll use to make our prediction rasters from the models we generate. This means we'll be cropping (or "clipping" in GIS terms) our prediction rasters. We'll then generate some background points to illustrate the partitioning methods. We'll sample 10,000 random points across one of our cropped rasters.

```{r, message=FALSE, fig.width=7, fig.height=5}
na.extent <- extent(c(-130,-50,25,45))  # rough bounding box for U.S. including most occurrence localities
preds.ext <- crop(preds, na.extent) #Crop predictor variables by this extent
backg <- randomPoints(preds.ext[[1]], n=10000) #Generate 10,000 random backround points for all Maxent runs
plot(preds.ext[[1]], ext=c(-100,-90,30,35))  # zoom in a bit to annual mean temp
points(backg)  # notice how we have pretty good coverage
```

We'll now take a look at the different partitioning methods offered by `ENMeval`.

```{r}

```

Now, we will perform model tuning using a variety of complexity combinations. Then we will isolate the results table of model tuning results, and using this table determine which model is the best approximation of fitting to the occurrence data without being overly complex. There are a variety of selection-techniques available, but for simplicity, here we will use AICc. We want the model with the lowest AICc score, therefore, we are interested in the model with a delta.AICc of 0. 

```{r}
models <- ENMevaluate(occ = species.coord, env = preds.ext, method='block', categoricals=NULL, 
                    fc = c("L", 'LQ', "H", "LQH"), bg.coords=backg, RMvalues=seq(1, 5, 0.5), parallel = T,
                    rasterPreds=T)
results.table<-models@results #Isolate the results table 
results.table #Look at the tuning results. Here, you will find information about omission rates, AUC and AICc scores
best.mod.AICc<- mod.table[mod.table[14]==0][c(1,16)] #Find the row of the results table that has a delta.AICc of 0.
```





## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
